\propertitle

### 2019-07-25 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-25T14:43:18+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

%\newcommand*{\yx}{\bm{\yxv}}%subpop state
%\newcommand*{\yxs}{\sav{\yx}}%subpop av state
%\newcommand*{\yX}{\bm{\yXv}}%pop state
%\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yaa}{\bm{\yav}}%subpop value
\newcommand*{\yaas}{\sav{\yaa\yaa}}%subpop av state
%\newcommand*{\yXXf}{\av{\yAA\yAA}}%subpop av state
\newcommand*{\ya}{\yav}%subpop av value
\newcommand*{\yAA}{\bm{\yAv}}%pop value
\newcommand*{\yA}{\yAv}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varIota_\textrm{u}}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
\newcommand*{\yI}{\varIota}
\newcommand*{\yD}{D}
%Lagr multipliers
\newcommand*{\yg}{\bm{c}}
\newcommand*{\yc}{\Hat{\bm{c}}}
\newcommand*{\yll}{\bm{\lambda}}
\newcommand*{\yl}{\bm{l}}
\newcommand*{\yk}{z}
\newcommand*{\yK}{\zeta}
%entropies
\newcommand*{\re}{\textrm{H}}
\newcommand*{\ybu}{H_{\text{B}}}

@@

Answering this kind of questions always engages an element of uncertainty.
We can't say \enquote{the answer is such and such}, but can at best assign
a degree of reasonable belief -- that is, probability -- to every possible
answer. The assessment of this distribution of belief involves experimental
data, such as recordings of neuronal activity from specific brain areas,
and pre-data knowledge about biological conditions and mechanisms. Our
pre-data \dobs\ are often simplified and called \enquote{models} to be
mathematically more tractable.

@@

The proposed model is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus. The maximum-entropy or minimum-relative-entropy method
\citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

We illustrate possible uses of our proposed model by applying it to two
concrete data sets:
\begin{enumerate*}[label=(\alph*)]
\item the activity of 65 neurons recorded for 20 min from a rat's Medial
  Entorhinal Cortex \citep{stensolaetal2012}, \item the activity of 159
  neurons recorded for 15 min from a macaque's Motor Cortex
  \citep{rostamietal2016_r2017}.
\end{enumerate*}
For each data set the model gives the most plausible frequency distribution
with which all levels of total activity of a larger network of 1\,000--10\,000
neurons appeared during the recording. These guessed frequency
distributions are distinctly different from the recorded ones for the
sampled neurons. For the first data set, for example, the frequency
distribution for the full network has two very distinct modes, both at low
activities, %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
see \fig~\ref{fig:plots_6mom}, whereas the frequency distribution for the
sample is monotonically decreasing with its maximum at zero activity. The
frequency distribution for the second data set doesn't have two modes but
still presents one prominent shoulder in its low-activity mode. Note that
these guessed features of the full network cannot be inferred by the
application of maximum-entropy \emph{to the sample alone}. Although we don't
formulate any hypotheses on why the larger network could have two distinct
low-activity regimes -- the purpose of the present work is only
methodological -- these results shows that the proposed method can lead to
the formulation or preliminary assessment of interesting hypotheses.

Maximum-entropy models have often been used to test whether a smaller part
of the sample data, for example time averages and second-order
correlations, is sufficient for quantifying our \dob\ about some aspects of
the neuronal activity of the sample. This use of maximum-entropy at the
sample level, however, has an important issue. We discuss this issue and
explain how our approach bypasses it. Moreover, our proposed model gives
clearer answers to questions of correlation sufficiency: we show this by
applying it to the two data set above addressing an example question of
sufficiency: two moments versus four moments.

Maximum-entropy models can be
considered as approximations of Bayesian models based on various
assumptions of inferential sufficiency
\citep{jaynes1986d_r1996,portamana2017}.
% Later on we'll show that our \dobs\ can be quantified by
% exclusively using the probability calculus. This derivation will provide a
% more accurate quantification, revealing that the maximum-entropy answer is
% only a first approximation.

We also apply the model to the question of informational sufficiency of
lower-order moments

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-25T15:46:11+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

%\newcommand*{\yx}{\bm{\yxv}}%subpop state
%\newcommand*{\yxs}{\sav{\yx}}%subpop av state
%\newcommand*{\yX}{\bm{\yXv}}%pop state
%\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yaa}{\bm{\yav}}%subpop value
\newcommand*{\yaas}{\sav{\yaa\yaa}}%subpop av state
%\newcommand*{\yXXf}{\av{\yAA\yAA}}%subpop av state
\newcommand*{\ya}{\yav}%subpop av value
\newcommand*{\yAA}{\bm{\yAv}}%pop value
\newcommand*{\yA}{\yAv}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varIota_\textrm{u}}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
\newcommand*{\yI}{\varIota}
\newcommand*{\yD}{D}
%Lagr multipliers
\newcommand*{\yg}{\bm{c}}
\newcommand*{\yc}{\Hat{\bm{c}}}
\newcommand*{\yll}{\bm{\lambda}}
\newcommand*{\yl}{\bm{l}}
\newcommand*{\yk}{z}
\newcommand*{\yK}{\zeta}
%entropies
\newcommand*{\re}{\Eta}
\newcommand*{\ybu}{\re_{\text{B}}}

@@

Answering this kind of questions always engages an element of uncertainty.
We can't say \enquote{the answer is such and such}, but can at best assign
a degree of reasonable belief -- that is, probability -- to every possible
answer. The assessment of this distribution of belief involves experimental
data, such as recordings of neuronal activity from specific brain areas,
and pre-data knowledge about biological conditions and mechanisms. Our
pre-data \dobs\ are often simplified to be mathematically more tractable,
and are therefore called \enquote{models}.

@@

The proposed model is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

We illustrate possible uses of our proposed model in
\sect~\ref{sec:application}, by applying it to two concrete data sets:
\begin{enumerate*}[label=(\alph*)]
\item the activity of 65 neurons recorded for 20 min from a rat's Medial
  Entorhinal Cortex \citep{stensolaetal2012}, \item the activity of 159
  neurons recorded for 15 min from a macaque's Motor Cortex
  \citep{rostamietal2016_r2017}.
\end{enumerate*}
For each data set the model gives the most plausible frequency distribution
with which all levels of total activity of a larger network of 1\,000--10\,000
neurons appeared during the recording. These guessed frequency
distributions are distinctly different from the recorded ones for the
sampled neurons. For the first data set, for example, the frequency
distribution for the full network has two very distinct modes, both at low
activities, %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
see \fig~\ref{fig:plots_6mom}, whereas the frequency distribution for the
sample is monotonically decreasing with its maximum at zero activity. The
frequency distribution for the second data set doesn't have two modes but
still presents one prominent shoulder in its low-activity mode. Note that
these guessed features of the full network cannot be inferred by the
application of maximum-entropy \emph{to the sample alone}. Although we don't
formulate any hypotheses on why the larger network could have two distinct
low-activity regimes -- the purpose of the present work is only
methodological -- these results shows that the proposed method can lead to
the formulation or preliminary assessment of interesting hypotheses.

Maximum-entropy models have often been used to test whether a smaller part
of the sample data, for example time averages and second-order
correlations, is sufficient for quantifying our \dob\ about some aspects of
the neuronal activity of the sample. This use of maximum-entropy at the
sample level, however, has an important issue. We discuss this issue and
explain how our approach bypasses it in \sect~\ref{sec:marginalization}.
Moreover, our proposed model gives clearer answers to questions of
correlation sufficiency: we show this by applying it to the two data set
above addressing an example question of sufficiency, two moments versus
four moments.

We obviously don't know whether the frequency distribution obtained with
our approach is the \emph{actual} one which the activity levels of the full
network had during the recording; it's only the most plausible. In high
dimensions, however, the features of the most plausible distribution may
\emph{not} be typical of the majority of most plausible distributions; and
the set of all possible frequency distributions, if the full network for
example comprises 1\,000 neurons, is a 1\,000-dimensional space. In
\sect~\ref{sec:assumptions} we therefore try to assess which features of
the frequency distribution delivered by our method may be typical and
therefore expected of the actual one. We find that general features such as
the bimodality of the first data set are indeed typical. Maximum-entropy
models can be considered as approximations of Bayesian models based on
various assumptions of inferential sufficiency
\citep{jaynes1986d_r1996,portamana2017}. How do our guesses change if we
modify our pre-data assumptions? We show, in the same section, that the
typical features indicated by our method are robust against such changes.

% (for a full network of 1\,000 neurons and
% a recording of 500\,000 time bins, for example, there are
% $\binom{500\,000+1\,000}{1\,000} \approx 10^{3\,131}$ possible frequency
% distributions).

% Later on we'll show that our \dobs\ can be quantified by
% exclusively using the probability calculus. This derivation will provide a
% more accurate quantification, revealing that the maximum-entropy answer is
% only a first approximation.
